Visualization and analysis of probability distributions of large temporal data

### 2021-11-06 Rmd/00-b-abstract.Rmd
putting reference even if unnumbered

# Abstract {-}

The research was motivated by the desire to understand the Australian smart meters data, which was collected half-hourly for two years at the household level. With temporal data available at ever finer scales, exploring periodicity can become overwhelming with so many possible temporal deconstructions to explore. Analysts are expected to comprehensively explore the many ways to view and consider temporal data. However, the plethora of choices and the lack of a systematic approach to do so quickly can make the task overwhelming.

<!-- exploration may be required to be carried out across both finer and coarser scales to draw useful inferences about the underlying process. It is important to be able to navigate through all the temporal deconstructions that accommodate for periodicities to have multiple perspectives of the observed data. This idea aligns with the notion of EDA (Tukey 1977) which emphasizes the use of multiple perspectives on data to help formulate hypotheses before proceeding to hypothesis testing. Slicing and dicing the data in all possible temporal scales as suggested by EDA can be a daunting task as it leads to a -->
<!-- myriad of possibilities. -->

<!-- Analysts are expected to comprehensively explore the many ways to view and consider temporal data. However, the plethora of choices and the lack of a systematic approach to do so quickly can make the task overwhelming -->

<!-- exploring periodicity can become overwhelming with so many possible temporal deconstructions to explore. Analysts are expected to comprehensively explore the many ways to view and consider temporal data. However, the plethora of choices and the lack of a systematic approach to do so quickly can make the task overwhelming. -->

@@

This work investigates how time may be dissected, resulting in alternative data segmentation and, as a result, different visualizations that can aid in the identification of underlying patterns. The first contribution (Chapter \@ref(ch:gravitas)) describes classes of time deconstructions using linear and cyclic time granularities. It provides tools to compute possible cyclic granularities from an ordered (usually temporal) index and also a framework to systematically explore the distribution of a univariate variable conditional on two cyclic time granularities by defining "harmony". "harmony" denotes pairs of granularities that could be analyzed together and reduces the search from all possible options. This approach is still overwhelming for human consumption because there would still be a huge number of harmonies. Hence, the second contribution (Chapter \@ref(ch:hakear)) refines the search of informative granularities by identifying those for which the differences between the displayed distributions are greatest and rating them in order of importance of capturing maximum variation. The third contribution (Chapter \@ref(ch:gracsr)) builds upon the first two to provide methods for exploring heterogeneities in repetitive behavior for many households and over multiple granularities. It accomplishes this by providing a way to cluster time series based on probability distributions across informative cyclic granularities. Although we were motivated by the smart meter example, the problem and the solutions we propose are practically relevant to any temporal data observed more than once per year.

<!-- there is a need to resort to clustering, which has been addressed in  It approaches the problem by computing pairwise distances between time series based on probability distributions over multiple cyclic granularities at once. -->
\newpage

>>>>

# Abstract {- #ch:abstract}

The research was motivated by the desire to understand the Australian smart meter data, which was collected half-hourly for two years at the household level. With temporal data available at ever finer scales,
exploration may be required to be carried out across both finer and coarser scales to draw useful inferences about the underlying process.
Analysts are expected to comprehensively explore the many ways to view and consider temporal data. However, the plethora of choices and the lack of a systematic approach to do so quickly can make the task daunting.

@@

This work investigates how time may be dissected, resulting in alternative data segmentation and, as a result, different visualizations that can aid in the identification of underlying patterns. The first contribution (Chapter \@ref(ch:gravitas)) describes classes of time deconstructions using linear and cyclic time granularities. It provides tools to compute possible cyclic granularities from an ordered (usually temporal) index and also a framework to systematically explore the distribution of a univariate variable conditional on two cyclic time granularities by defining "harmony". "harmony" denotes pairs of granularities that could be effectively analyzed together and reduces the search from all possible options. This approach is still overwhelming for human consumption because there would still be a huge number of harmonies. The second contribution (Chapter \@ref(ch:hakear)) refines the search for informative granularities by identifying those for which the differences between the displayed distributions are greatest and also rating them in order of importance of capturing maximum variation. The third contribution (Chapter \@ref(ch:gracsr)) builds upon the first two to provide methods for exploring heterogeneities in repetitive behavior for many time series and over multiple granularities. It accomplishes this by providing a way to cluster time series based on probability distributions across informative cyclic granularities. Although we were motivated by the smart meter example, the problem and the solutions we propose are practically relevant to any temporal data observed more than once per year.

\newpage

### 2021-11-06 Rmd/00-d-acknowlegements.Rmd

 <!-- - PhD friends and R community -->
Thank you, Stephanie and Nicholas Tierney, for being so warm, kinds and welcoming when I began my PhD. Being a part of such an inclusive environment meant a lot to me as an international student. Thank you, Nicholas Spyrison and Mitchell O'Hara-Wild, for the numerous informative discussions, brainstorming sessions, as well as for generously sharing your wisdom and experience in software development. Thank you to each of you, NUMBATs, and the broader R community for directly and indirectly helping me countless times. You never fail to inspire me.

@@

<!-- add people who are helping you with proofreading -->

>>>>

 <!-- - PhD friends and R community -->
Thank you, Stephanie and Nicholas Tierney, for being so warm, kind and welcoming when I began my PhD. Being a part of such an inclusive environment meant a lot to me as an international student. Thank you, Nicholas Spyrison and Mitchell O'Hara-Wild, for the numerous informative discussions, brainstorming sessions, as well as for generously sharing your expertise and experience in software development. Thank you to each of you, NUMBATs, and the broader R community for directly and indirectly helping me countless times. You never fail to inspire me.

@@

<!-- add people who are helping you with proofreading -->
<!-- Add Peter -->
<!-- Although the journey is coming to an end, you wish to take these life learnings forward -->
<!-- add Earo and Stuart that you followed their model of workflow and templates, using the power of reproducible and accessible  research -->

### 2021-11-06 Rmd/05-conclusion.Rmd

The **gravitas** package provides very general tools to compute and manipulate cyclic granularities and generate plots displaying distributions conditional on those granularities. The functions `search_gran()`, `create_gran()`, `harmony()`, `gran_advice()` and `prob_plot()` provides a workflow for an analyst to systematically explore large quantities of temporal data across different harmonies (pairs of granularities that can be analyzed together.) This package was developed as part of my internship at Google Summer of Code, 2019. It and has been on CRAN since January 2020. The website (<https://sayani07.github.io/gravitas>) includes full documentation and two vignettes about the package usage. There has been a total of 12K downloads from the RStudio mirror dating from 2020-11-01 to 2021-11-01. This package supplements the paper corresponding to \@ref(ch:gravitas), which has won the ACEMS Business Analytics Award 2021. The package can be generalized to non-temporal applications for which a hierarchical structure can be construed similar to time.

@@

The R package hakear (https://github.com/Sayani07/hakear) implements ideas presented in Chapter @ref (ch:hakear). The function wpd() computes the weighted pairwise distances ($wpd$) for each cyclic granularity or pair of granularities, and select harmonies() selects the ones with significant patterns and ranks them from highest to lowest $wpd$. These selected harmonies can be plotted using package `gravitas` for potentially interesting displays.
Because $wpd$ is computationally expensive, this package uses the R package `parallel` to handle parallel computations.

@@

The open-source R package `gracsr` is available on Github (https://github.com/Sayani07/gracsr) to implement ideas presented in Chapter \@ref(ch:gracsr). The package contains functions for implementing the complete clustering methodology with choices of scaling and distance metrics discussed in the paper. It has received an AUD 3000 grant as part of the ACEMS Business Analytics Prize towards polishing the functions and preparing it for CRAN.

@@

I plan to integrate `hakear` with `gravitas` as one R package to systematically exploring few time series, whereas `gracsr` would provide the clustering framework for exploring many time series together. I plan to run it through https://ropensci.org/software-review/ to develop them further for more visibility and efficient usage.

@@

With the volume of data projected to grow in the future, potentially leading to increased variability in patterns, research is needed to understand the issues that may arise with the existing methodology and how to adapt our current algorithm. Computational efficiency is also critical when scaling up for analysis of huge data sets, let alone when adding features to existing highly dimensional data structures.

Testing needs to be carried out with non-hierarchy based clustering methods to see whether these distances perform better with other algorithms. Also, to evaluate how much information is lost when aggregating individual level demand versus distributions.

@@

We provide solutions that are realistically applicable to any temporal data observed more than once per year. We could just verify its value in the electricity smart meter context. With numerous open-source benchmark data sets accessible, it is necessary to test how well the approaches operate in various disciplines.

## Final words

>>>>

The **gravitas** package provides very general tools to compute and manipulate cyclic granularities and generate plots displaying distributions conditional on those granularities. The functions `search_gran()`, `create_gran()`, `harmony()`, `gran_advice()` and `prob_plot()` provides the entire workflow for an analyst to systematically explore large quantities of temporal data across different harmonies (pairs of granularities that can be analyzed together.) This package was developed as part of my internship at Google Summer of Code, 2019. It has been on CRAN since January 2020. The website (<https://sayani07.github.io/gravitas>) includes full documentation and two vignettes about the package usage. There has been a total of 12K downloads from the RStudio mirror dating from 2020-11-01 to 2021-11-01. This package supplements the paper corresponding to Chapter \@ref(ch:gravitas), which has won the ACEMS Business Analytics Award 2021. The package can be generalized to non-temporal applications for which a hierarchical structure can be construed similar to time.

@@

The R package hakear (https://github.com/Sayani07/hakear) implements ideas presented in Chapter \@ref(ch:hakear). The function wpd() computes the weighted pairwise distances ($wpd$) for each cyclic granularity or pair of granularities, and `select_harmonies()` selects the ones with significant patterns and ranks them from highest to lowest $wpd$. These selected harmonies can be plotted using package `gravitas` for potentially interesting displays. This package is reliant on parallel processing using multiple multi-core computers for faster computation of $wpd$.

@@

The open-source R package `gracsr` is available on Github (https://github.com/Sayani07/gracsr) to implement ideas presented in Chapter \@ref(ch:gracsr). The package contains functions for implementing the complete clustering methodology with choices of scaling and distance metrics discussed in the chapter. It has received a grant (AUD 3000) as part of the ACEMS Business Analytics Prize towards polishing the functions and preparing it for CRAN.

@@

I want to combine 'hakear' and 'gravitas' into a single R package for systematically exploring distributions of a smaller number of time series, whilst 'gracsr' would offer the clustering framework for systematically exploring a large number of time series. I intend to run it via https://ropensci.org/software-review/ to further refine them for more exposure and efficiency.

### Scaling up the methods for multivariate time series data

With multiple time series available for each observation, the complexity of efficient exploration and visualization grows exponentially. In this case, conditional distributions include not only temporal dependency but also variables and their dependencies. This adds to the already high-dimensional data structures that result from studying distributions. Current research on visualizing multivariate data must account for time's inherent characteristics in order to efficiently explore distributions of multivariate time series data.

Testing needs to be carried out with non-hierarchy based clustering methods to see whether the proposed distances perform better with other algorithms. It would also be helpful to know how much more value we obtain from considering distributions rather than aggregating time series data. In the case of smart meters, this could be accomplished by comparing the results while validating them against external data such as weather conditions, socioeconomic or other demographic factors of those households.

@@

With the volume of data expected to increase in the future, potentially leading to increased variability in patterns, more experiments are needed to understand what sort of problems may arise with our current methodology and how to improve our current algorithm to address those issues. All of this must be done while keeping computational efficiency in mind, as this is critical when scaling up for analysis of large data sets.

### Develop inferential methods for the clusters

It is crucial to determine whether or not any detected clusters or patterns are statistically significant. In Chapter \@ref(ch:gracsr), we have used the permutation approach to detect if patterns are authentic or not across one or a pair of granularities. A similar method for finding significance across clusters is not appropriate if the observations are assumed to be independent. Most large temporal data sets gathered at fine temporal scales usually include several repetitive patterns. Therefore, the permutations used for comparisons must account that different temporal granularities have distinct conditional distributions.

@@

We propose solutions that are realistically applicable to any temporal data collected more than once per year and may be implemented in a variety of settings. We could simply verify its utility in the context of smart meter electricity data. With the availability of several open-source benchmark data sets, it is vital to evaluate how well the methodologies perform across a variety of fields.
